Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-12 manuscript_sciad.Rmd
edits: resubmit to sciad

---
title: "Title: Ecosystem size and complexity dictate riverine biodiversity"
output:
  pdf_document:
    latex_engine: xelatex
csl: science.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

**One sentence summary:** This study provides the first evidence that ecosystem size and complexity play comparable roles in regulating biodiversity.

@@

**Abstract:** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in scaling biodiversity is largely unknown. Here, we use theory and extensive fish community data from two distinct geographic regions (Japan and United States) to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual control of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

# Maintext

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how ecosystem complexity controls biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

## Synthesizing ecosystem size and complexity influences

@@

## Empirical evidence from distinct geographic regions

@@

## Implications for biodiversity conservation

@@

While the prevailing evidence supports the importance of ecosystem size in scaling biodiversity patterns [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings of the dual control of biodiversity should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
output:
  pdf_document:
    latex_engine: xelatex
csl: science.csl
knit: (function(inputFile, encoding) {
        rmarkdown::render(inputFile, encoding = encoding, output_dir = "document_output")
      })
bibliography: references.bib
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
---

@@

@@

**Abstract:** Larger ecosystems support more species; this ubiquitous pattern is the foundation of current conservation schemes. However, many ecosystems possess a complex spatial structure that cannot be represented by area, and the role of such complexity in controlling biodiversity is largely unknown. Here, we use theory and extensive fish community data from two distinct geographic regions (Japan and United States) to show that ecosystem size and complexity dictate riverine biodiversity. We found that larger and more branched 'complex' river networks harbored greater species richness due to increased space and environmental heterogeneity. The complexity effect was comparable to the size effect, and this pattern has emerged regardless of ecological contexts. The dual control of biodiversity may be a pervasive feature that has far-reaching implications for biodiversity conservation.

## Introduction

@@

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity [@turnerLandscapeEcologyTheory2015; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few [@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike [@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms [@grantLivingBranchesPopulation2007]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure [@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @brownDendriticNetworkStructure2010], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems [@rinaldoRiverNetworksEcological2018]. For example, these studies have highlighted key roles of branching structure in driving local biodiversity patterns, such as increased species richness at merging points of branches [@carraraDendriticConnectivityControls2012]. However, most empirical research has explored the consequences of branching complexity for local community structure [@erosLandscapeEcologyRivers2019] or has relied solely on theoretical arguments with limited replications of branching architecture [@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity controls biodiversity patterns at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where accelerated species loss threatens the delivery of ecosystem services [@reidEmergingThreatsPersistent2019].

@@

## Results and Discussion

### Theoretical synthesis of ecosystem size and complexity influences

@@

### Empirical evidence from distinct geographic regions

@@

### Implications

@@

While the prevailing evidence supports the importance of ecosystem size in scaling biodiversity patterns [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.
